From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-02-11 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Update classical-phil-article.tex
incorporated edits from Monday 2-8 meeting to first half of paper.

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Despite being coordinate-invariant, the density itself is coordinate-dependent. We therefore argue that some quantities are physically significant despite being coordinate-dependent. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we consider both the gauge redundancy of the conjugate momentum (Section~\ref{fundamental}) and of the Lagrangian (Section~\ref{gauge}).

@@

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy the condition of motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space. Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.
% [To do: verify that the scenario of two different Lagrangians---giving rise to the same trajectories in physical space with different energies---arises ONLY when the motion equivalence condition fails. See Curiel 2014, page 42 for his take on this kind of scenario.]

Of course, understanding the relationship between Hamiltonian and Lagrangian mechanics partly depends on how one defines ``classical systems." Different definitions can result in different interpretive relationships. On our account, Newtonian systems are the widest domain of classical systems, including any system that is infinitesimally reducible. Hamiltonian mechanics picks out the subset of Newtonian systems that follow deterministic and reversible evolution. Our treatment thereby sets aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative, i.e. conserving energy. Consequently, our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although the Lagrangian formulation applies to dissipative systems \parencites[]{Smith,Curiel}, this is not what we will mean by ``Lagrangian systems."\footnote{Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view non-dissipative systems as Newtonian, but neither Lagrangian nor Hamiltonian. Before presenting the derivational details of our argument in Section~\ref{derivation}, we clarify what we mean by ``fundamental."

[XXX To do: Clarify the scope of our treatment: our discussion makes it seem restricted to $n$-particle classical mechanics. Does it also handle rigid bodies and classical continuum mechanics? Cite Mark Wilson \parencite*[]{Wilson} on these issues].

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the ``real" fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure as part of the theory's ontology, but as a formal device for encoding physical information or as a calculational tool.} Whether Hamiltonian phase space structure is metaphysically fundamental (in a classical world) does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically more fundamental, and we will argue for its epistemic and conceptual primacy as well. Indeed, our argument does not privilege any one of the at least three different mathematical ways of formalizing Hamiltonian mechanics [XXX to do: mention these different Hamiltonian formalisms and provide references]. Each of these frameworks has different pragmatic costs and benefits. Hence, it is plausible that there may be no single, uniquely necessary mathematical structure for mechanics, but rather a family of related structures. In other words, there may be no canonical minimal structure for formalizing the relevant physical models and physical assumptions.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics supplies a further reason for viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry" \parencites*[46]{Teh}.}  Additionally, each of these formalisms references some ``unphysical" coordinates or representational means, whose physical significance is borne out by their fruitfulness in modeling. We return to the physical significance of certain coordinate choices in Section~\ref{fundamental}.
%The mathematical structure that you ``need" might not be unique. Neither one of these three formalizations of Hamiltonian mechanics is necessarily more ``fundamental" or joint-carving than the others. Nevertheless, there are compelling physical and pragmatic motivations for preferring the extended phase space formulation of Hamiltonian mechanics. In this formulation, we gain an invariant Hamiltonian that is the conjugate variable of the proper time, while the normal Hamiltonian (which is not coordinate-invariant) still plays the role of generating time-translations.

@@

This section justifies the physical and mathematical relationships in Figure~\ref{diagram}.
Sections~\ref{infinitesimal}-\ref{deterministic} show how to derive the formalism of Hamiltonian mechanics from the assumptions of infinitesimal reducibility and deterministic/reversible evolution. Section~\ref{motion} shows how systems that satisfy the further condition of motion equivalence can be treated within the formalism of Lagrangian mechanics. We thus demonstrate in detail our methodology of deriving  relevant mathematical structures from underlying physical conditions.

@@

Let $\mathcal{C}$ be the state space for the whole system, so that any point $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point in $\mathcal{S}$ describes a possible state of a classical particle. We assume that this state space $\mathcal{S}$ is a manifold.\footnote{Ultimately, we derive an if-then claim: if the state space of classical particles is a manifold, then it must be a symplectic manifold.} Then for each state $c \in \mathcal{C}$ there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state. Formally, if $\mathcal{S}$ is a manifold, $\int_U \rho_c d\mathcal{S}$ gives us the fraction of the system whose parts are in the region $U$ of the state space $\mathcal{S}$. 
%[is region U in the state space C or S. I assume in S.].

We define the \textit{state variables} as a set of quantities $\xi^a$ that fully identify a state. Formally, $\xi^a: \mathcal{S} \to \mathbb{R}^n$; this function yields a real-valued vector for any given particle state, possessing $n$-many degrees of freedom.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates,' but we will reserve that terminology for the canonical position coordinates, which are the state variables that define physical units.} We will define a \textit{coordinate} $q \in \xi^a$ as a particular state variable that defines a unit.  For instance, $q$ might be position, defining units of meters. In general, the coordinates are the subset of the state variables that define the unit system, needed to measure properties of the system.

% % Ask GC if instances of `state variables' in the next paragraph are correct; i.e. `state variables' rather than `coordinates'

On physical grounds, the distribution $\rho_c$ must transform as both a density and a scalar function. As a function defined over the state variables, the distribution transforms as a density, i.e. $\rho_c$ changes as a function of the Jacobian determinant that relates new state variables to old ones. As a function defined over the points in phase space, the distribution transforms as a scalar, i.e. $\rho_c$ is invariant under all state variable transformations. Hence, the Jacobian determinant must in fact equal one. This means that $\rho_c$ is a function only of the particle state $s$ and does not depend on the choice of coordinates.\footnote{Formally, we can express this argument as a chain of equalities: $\rho_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho_c(\hat\xi^b)$, where $\rho_c(\xi^a)$ and $\rho_c(\hat{\xi}^b)$ represent the same distribution as expressed over different state variables and $s(\xi^a)$ and $s(\hat{\xi}^b)$ represent the same particle state as expressed by the different variables. Note that $\rho_c(s(\hat{\xi}^b))$ shows how the distribution depends on the state variables only through the particle state.} We will now show that in order for the distribution to satisfy these two constraints, the classical phase space must be at least a symplectic manifold. In other words, having symplectic manifold structure is a necessary condition for the distribution over particle states to transform as a density and scalar function.
%This is because the distribution is a function only of the states, rather than of the particular choice of coordinates. 
 % Classical phase space (i.e. a symplectic manifold) is the only space that satisfies these constraints on the distribution. 
%(in manifold terminology, these are known as coordinates)

We begin with the simplest case: suppose that a single coordinate $q $ suffices to define the unit system for the particle state space. Then, the state variables are given by a set $\xi^a = \{ q, k^i \}$, where \textit{prima facie} the index $i$ can range over more than one number. We will show that, in fact, there can only be a single $k^i$ variable in this case, conjugate to the single $q$ coordinate. Assuming that the particle density $\rho_c$ is a scalar, it follows that we can arbitrarily change the unit defined by $q$ without changing the density $\rho_c$. 
% We further assume that $q$ fully defines the units for all state variables. [Is there a way to physically motivate this assumption, beyond this being a constitutive role of the position-like state variables?]

Now, suppose that we change the units by transforming the coordinate $q$ to a new coordinate $\hat{q}$. Since---by assumption---$q$ fully defines the unit system, $\hat{q}$ must do so as well, making it a function just of the old coordinate: $\hat{q}=\hat{q}(q)$. Call the new state variables $\hat{\xi}^b = \{ \hat{q}, \hat{k}^i\}$. As shown above, the Jacobian determinant $\left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$ must equal one. Assume for contradiction that there is only one state variable, so that there are no $k^i$ variables. Then we would have $\left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| = \left|\frac{\partial \hat q}{\partial q} \right| = 1$. But this would mean the unit change cannot be arbitrary. Therefore we must have two or more state variables, so that there must be at least one $k^i$ variable.

@@

This equation provides only a single constraint, placed on the determinant of the transformation from $k^a$ to $\hat{k}^b$. Hence, if there were three or more $\xi^a$ variables, this constraint would be insufficient to recover the transformation of the $k^a$ uniquely. Yet in this case, $\hat q$  would not fully define the units for all other state variables, which is contrary to assumption. Consequently, there must be exactly two state variables: $q$ and a single $k$. So in this case of a single $q$ variable, coordinate-independent areas and densities can only be defined on a two-dimensional manifold.

@@

To generalize, we say two coordinates are \textit{independent} if changing the units for one does not change the units for the other. Suppose our particle state space $\mathcal{S}$ is such that its units are fully defined by $n$ independent coordinates $q^i$. If we change the first coordinate $q^1$ without changing the others, then the variable $k_1$ must change as before, so that the density remains invariant. Likewise, transforming the second coordinate $q^2$ in the same way (while fixing $k_1$) requires a corresponding $k_2$. In the end, we will find that $\mathcal{S}$ is $2n$-dimensional, and the state variables are $\xi^a = \{ q^i, k_i \}$. We define an \textit{independent degree of freedom} as the space charted by a pair of such variables.

@@

The conclusion is that, under the infinitesimal reducibility assumption, a state $c$ of the system is a coordinate-invariant distribution over the state space $\mathcal{S}$ of its infinitesimal parts. If the state space of the particles is a manifold, then it must be a symplectic manifold: symplectic manifold structure is necessary for the distribution to be coordinate-invariant. 
% Could connect this with the earlier discussion of North's methodology of seeking coordinate-invariants, from section~\ref{fundamentality}.

@@

If we focus on the parts of the system, this means that for every evolution in phase space there should be one and only one trajectory in physical space. Note that each space variable $x^i$ is a coordinate, i.e. a state variable that defines a unit. In fact, the trajectories can be fully described by these units and only these units. For, to specify a trajectory, we need epistemic access to a unit system, where the number of independent units is determined by the dimensionality of physical space. It follows that $x^i = q^i$, each $q^i$ is paired with a conjugate $p_i$, and each state $\{q^i, p_i\}$ corresponds to one and only one trajectory in physical space. At each point $x^i$, then, there must be infinitely many trajectories, one for each possible combination of $\{p_i\}$. Since these trajectories are differentiable in $x^i$, we can define a velocity $v^i = d_t x^i$. If the equations of motions were such that $v^i=v^i(q^i)$, then motion equivalence would fail as the full trajectory would be determined only by the $q^i$. Hence, we must have $v^i=v^i(q^j, p_k)$, so that knowledge of the state in phase space suffices for knowledge of the trajectory. If motion equivalence holds, then this relationship must be invertible, meaning that knowledge of the trajectory suffices for knowledge of the phase space state. For systems that satisfy the motion equivalence condition, then, the following relationships hold at any given time:
\begin{equation}\label{weak_equivalence}
\begin{aligned}
x^i &= q^i \\
v^i &= d_t x^i = v^i(q^j, p_k)
\end{aligned}
\end{equation}

In this case---which we call \textit{weak equivalence}---$v^i(q^j, p_k)$ is invertible at every $q^i$. Therefore, for fixed q, the Jacobian matrix  $\frac{\partial v^i}{\partial p_j}$ must be invertible. Since $v^i = d_t q^i = \frac{\partial H}{\partial p_i}$, the Hessian $\frac{\partial^2 H}{\partial p_i \partial p_j} = \frac{\partial}{\partial p_i} (\frac{\partial H}{\partial p_j})$ equals the Jacobian $\frac{\partial v^i}{\partial p_j}$. Since the Jacobian is invertible, the Hessian must be non-zero everywhere, and therefore it must have the same sign everywhere, which we take to be positive. These are exactly the cases where a Lagrangian can be constructed, and where the Lagrangian leads to a unique solution (via the stationary action principle). Hence, weak equivalence suffices for the existence of a hyper-regular Lagrangian.\footnote{Weak equivalence corresponds to a general treatment of Lagrangian mechanics that does not presuppose Riemannian geometry. As we discuss in Section~\ref{Curiel}, Equations~\ref{weak_equivalence} correspond to the kinematical constraints central to Curiel's \parencites*[]{Curiel} argument for privileging Lagrangian mechanics.}

>>>>

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Despite being coordinate-invariant, one cannot make the density truly coordinate independent. We therefore argue that some quantities are physically significant despite being coordinate-dependent. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we consider both the gauge redundancy of the conjugate momentum (Section~\ref{fundamental}) and of the Lagrangian (Section~\ref{gauge}).

@@

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy the condition of motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space.\footnote{Since photons travel at constant speed, we can only recover the direction of momentum from the trajectory, but not the magnitude. Two photons with different momenta can follow the same trajectory.} Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.
% [To do: verify that the scenario of two different Lagrangians---giving rise to the same trajectories in physical space with different energies---arises ONLY when the motion equivalence condition fails. See Curiel 2014, page 42 for his take on this kind of scenario.]

Of course, different definitions of ``Hamiltonian" and ``Lagrangian systems" can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems satisfy the motion equivalence assumption. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they constitute the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems \parencites[\S 10.4]{Cline}, we exclude these.\footnote{Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications. For details of the Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view non-dissipative systems as neither Lagrangian nor Hamiltonian. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to continuum mechanics (including rigid bodies and fluid mechanics).\footnote{For discussion of continuum mechanics in the context of classical mechanics, see \textcites[]{Wilson}.} Before presenting the derivational details of our argument in Section~\ref{derivation}, we clarify what we mean by ``fundamental."
% For fluid systems, the deterministic/reversible evolution condition applies to the whole system but not for the individual particles due to interactions between these.}
% % Why don't interactions between particles also arise for $n$-particle mechanics? e.g. Newtonian gravity?

% % XXX: try to find a different citation for the claim that dissipative forces can be incorporated in both Hamiltonian and Lagrangian mechanics. Cline libre text might not look the best!

% https://phys.libretexts.org/Bookshelves/Classical_Mechanics/Book%3A_Variational_Principles_in_Classical_Mechanics_(Cline)/10%3A_Nonconservative_Systems/10.04%3A_Rayleighs_Dissipation_Function

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the ``real" fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure as part of the theory's ontology, but as a formal device for encoding physical information or as a calculational tool.} Whether Hamiltonian phase space structure is metaphysically fundamental (in a classical world) does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically more fundamental, and we will argue for its epistemic and conceptual primacy as well. Indeed, our argument does not privilege any one of the various mathematical formalizations of Hamiltonian mechanics.\footnote{These include formulating Hamiltonian mechanics on a $2n+1$-dimensional contact manifold, on an extended phase space, on a Poisson manifold, and the standard formulation on a symplectic manifold that we typically use below.} Each of these frameworks has different pragmatic costs and benefits. Hence, it is plausible that there may be no single, uniquely necessary mathematical structure for mechanics, but rather a family of related structures. In other words, there may be no canonical minimal structure for formalizing the relevant physical models and physical assumptions.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics supplies a further reason for viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry" \parencites*[46]{Teh}.}  Additionally, each of these formalisms references some ``unphysical" coordinates or representational means, whose physical significance is borne out by their fruitfulness in modeling. We return to the physical significance of certain coordinate choices in Section~\ref{fundamental}.
%The mathematical structure that you ``need" might not be unique. Neither one of these three formalizations of Hamiltonian mechanics is necessarily more ``fundamental" or joint-carving than the others. Nevertheless, there are compelling physical and pragmatic motivations for preferring the extended phase space formulation of Hamiltonian mechanics. In this formulation, we gain an invariant Hamiltonian that is the conjugate variable of the proper time, while the normal Hamiltonian (which is not coordinate-invariant) still plays the role of generating time-translations.

@@

This section justifies the physical and mathematical relationships in Figure~\ref{diagram}.
Sections~\ref{infinitesimal}-\ref{deterministic} derive the formalism of Hamiltonian mechanics from the assumptions of infinitesimal reducibility and deterministic/reversible evolution. Section~\ref{motion} shows how systems that satisfy the further condition of motion equivalence can be treated within the formalism of Lagrangian mechanics. We thus demonstrate in detail our methodology of deriving  relevant mathematical structures from underlying physical conditions.

@@

Let $\mathcal{C}$ be the state space for the whole system, so that any point $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point in $\mathcal{S}$ describes a possible state of a classical particle. It can be shown that this state space must be a manifold, given the continuous evolution in time of the state space variables.\footnote{For details of this derivation, see \textcites[]{AoPPhy1} [omit this for review; or find alternative source].} Then for each state $c \in \mathcal{C}$ there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state. Formally, if $\mathcal{S}$ is a manifold, $\int_U \rho_c d\mathcal{S}$ gives us the fraction of the system whose parts are in the region $U$ of the state space $\mathcal{S}$. 
%[is region U in the state space C or S. I assume in S.].
% % XXX could try to find justification/derivation for the state space being a manifold in the Kevin Kelly book, or could cite the math paper with Greenfield; proposition 6.2. Try to include a better gloss/motivation for this claim (basic idea: if time is continuous, trajectories in state space must be continuous, local structure must be compatible). Curiel paper might also have a brief justification/motivation, if I recall

We define \textit{state variables} as a set of quantities $\xi^a$ that fully identify a state. Formally, $\xi^a: \mathcal{S} \to \mathbb{R}^n$; this function yields a real-valued vector for any given particle state, possessing $n$-many degrees of freedom.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates', terminology that we adopt when speaking about coordinate transformations, coordinate invariance, and coordinate independence.} We will define a \textit{unit variable} $q \in \xi^a$ as a particular state variable that defines a unit.  For instance, $q$ might be position, defined in units of meters. In general, the unit variables are the subset of the state variables that define the unit system, which is needed to measure properties of the system. 
% [XXX: coordinates -> unit variables]

% % Ask GC if instances of `state variables' in the next paragraph are correct; i.e. `state variables' rather than `coordinates'
% On physical grounds, [XXX: I'd say it's more mathematical grounds]

Importantly, the distribution $\rho_c$ must transform as both a density and a scalar function. As a density, $\rho_c$ is defined over the state variables $\xi^a$, so that we can take integrals of the form $\int_U \rho_c(\xi^a) d\xi^a$ (in this context, $\rho_c:  \mathbb{R}^n \to \mathbb{R}$). The density $\rho_c(\xi^a)$ changes as a function of the Jacobian determinant that relates new state variables to old ones. As a function defined over the points in phase space ($\rho_c : \mathcal{S} \to \mathbb{R} $), the distribution transforms as a scalar. Hence, $\rho_c$ is invariant under all state variable transformations. Consequently, the Jacobian determinant must in fact equal one. This means that $\rho_c$ is a function only of the particle state $s$ and is invariant under changes of units.\footnote{Formally, we can express this argument as a chain of equalities: $\rho_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho_c(\hat\xi^b)$, where $\rho_c(\xi^a)$ and $\rho_c(\hat{\xi}^b)$ represent the same distribution as expressed over different state variables and $s(\xi^a)$ and $s(\hat{\xi}^b)$ represent the same particle state as expressed by the different variables. Note that $\rho_c(s(\hat{\xi}^b))$ shows how the distribution depends on the state variables only through the particle state.} We will now show that in order for the distribution to satisfy these two constraints, the classical phase space must be at least a symplectic manifold. In other words, having symplectic manifold structure is a necessary condition for the distribution over particle states to transform as a density and scalar function.
%This is because the distribution is a function only of the states, rather than of the particular choice of coordinates. 
 % Classical phase space (i.e. a symplectic manifold) is the only space that satisfies these constraints on the distribution. 
%(in manifold terminology, these are known as coordinates)

We begin with the simplest case: suppose that a single unit variable $q $ suffices to define the unit system for the particle state space. Then, the state variables are given by a set $\xi^a = \{ q, k^i \}$, where \textit{prima facie} the index $i$ could be any non-negative integer. We will show that, in fact, there must be exactly one $k^i$ variable, conjugate to the single $q$ variable. 
% We further assume that $q$ fully defines the units for all state variables. [Is there a way to physically motivate this assumption, beyond this being a constitutive role of the position-like state variables?]

Since the particle density $\rho_c$ is a scalar, we can arbitrarily change the unit defined by $q$ without changing the density $\rho_c$. So suppose we change the units by transforming $q$ to a new unit variable $\hat{q}$. Since---by assumption---$q$ fully defines the unit system, $\hat{q}$ must do so as well, making it a function of just the old unit: $\hat{q}=\hat{q}(q)$. Call the new state variables $\hat{\xi}^b = \{ \hat{q}, \hat{k}^i\}$. As shown above, the Jacobian determinant $\left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$ must equal one. Assume for contradiction that there is only one state variable, so that there are no $k^i$ variables. Then we would have $\left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| = \left|\frac{\partial \hat q}{\partial q} \right| = 1$. But this would mean the unit change cannot be arbitrary. Therefore we must have two or more state variables, requiring at least one $k^i$ variable.

@@

This equation provides only a single constraint, placed on the determinant of the transformation from $k^a$ to $\hat{k}^b$. Hence, if there were three or more $\xi^a$ variables, this constraint would be insufficient to recover the transformation of the $k^a$ uniquely. Yet in this case, $\hat q$  would not fully define the units for all other state variables, contrary to assumption. Consequently, there must be exactly two state variables: $q$ and a single $k$. So in the case of a single $q$ variable, defining a coordinate-independent density requires a two-dimensional manifold.

@@

Extending to the general case, note that two unit-variables are \textit{independent} if changing the units for one does not change the units for the other. Suppose our particle state space $\mathcal{S}$ is such that its units are fully defined by $n$ independent unit-variables $q^i$. If we change the first unit-variable $q^1$ without changing the others, then the variable $k_1$ must change as before, so that the density remains invariant. Likewise, transforming the second unit-variable $q^2$ in the same way (while fixing $k_1$) requires a corresponding $k_2$. In the end, we will find that $\mathcal{S}$ is $2n$-dimensional, and the state variables are $\xi^a = \{ q^i, k_i \}$. We define an \textit{independent degree of freedom} as the space charted by a pair of such variables.

@@

The conclusion is that, under the infinitesimal reducibility assumption, a state $c$ of the system is a coordinate-invariant distribution over the state space $\mathcal{S}$ of its infinitesimal parts. We have shown that the state space of the particles must be a symplectic manifold. Additionally, any symplectic manifold structure supports the definition of coordinate-invariant distributions: any scalar function can be integrated using the symplectic form in a coordinate-invariant way. Therefore the symplectic structure is exactly the structure needed to define an invariant density. Section~\ref{density} describes further interpretive consequences of this connection between invariant densities and symplectic structure.
% % [XXX: clarify/clean-up this ending. also, emphasize/foreshadow the physical signifiance of the distribution and density]. also, clarify relationship b/w distribution (ontology) and density (epistemic access to/representation of this ontology)
%Symplectic manifold structure is exactly the structure needed to define an invariant density. 
% Could connect this with the earlier discussion of North's methodology of seeking coordinate-invariants, from section~\ref{fundamentality}.

@@

If we focus on the parts of the system, this means that for every evolution in phase space there should be one and only one trajectory in physical space. Note that each space variable $x^i$ is a unit variable, i.e. a state variable that defines a unit. In fact, the trajectories can be fully described only by these unit variables. For, to specify a trajectory, we need epistemic access to a unit system, where the number of independent units is determined by the dimensionality of physical space. It follows that (i) $x^i = q^i$, (ii) each $q^i$ is paired with a conjugate $p_i$, and (iii) each state $\{q^i, p_i\}$ corresponds to one and only one trajectory in physical space. At each point $x^i$, then, there must be infinitely many trajectories, one for each possible combination of $\{p_i\}$. Since these trajectories are differentiable in $x^i$, we can define a velocity $v^i = d_t x^i$. If the equations of motion were such that the velocity were a function of only the unit variables (i.e. $v^i=v^i(q^i)$), then motion equivalence would fail. For if both $x^i$ and $v^i$ were a function of only $q^i$, this relationship would not be invertible because of dimensional considerations (a surjective function from an $n$-dimensional manifold to a $2n$-dimensional manifold cannot be invertible). Hence, we must have $v^i=v^i(q^j, p_k)$, so that knowledge of the state in phase space suffices for knowledge of the trajectory. If motion equivalence holds, then this relationship must be invertible, meaning that knowledge of the trajectory suffices for knowledge of the phase space state. For systems that satisfy the motion equivalence condition, then, the following relationships hold at any given time:
\begin{equation}\label{weak_equivalence}
\begin{aligned}
x^i &= q^i \\
v^i &= d_t x^i = v^i(q^j, p_k)
\end{aligned}
\end{equation}

In this case---which we call \textit{weak equivalence}---$v^i(q^j, p_k)$ is invertible at every $q^i$. Therefore, for fixed $q$, the Jacobian matrix  $\frac{\partial v^i}{\partial p_j}$ must be invertible. Since $v^i = d_t q^i = \frac{\partial H}{\partial p_i}$, the Hessian $\frac{\partial^2 H}{\partial p_i \partial p_j} = \frac{\partial}{\partial p_i} (\frac{\partial H}{\partial p_j})$ equals the Jacobian $\frac{\partial v^i}{\partial p_j}$. Since the Jacobian is invertible, the Hessian must be non-zero everywhere, and therefore it must have the same sign everywhere, which we take to be positive. Weak equivalence allows us to express particle states in terms of either position/momentum or position/velocity variables. These are exactly the cases where a Lagrangian can be constructed, and where the Lagrangian leads to a unique solution (via the stationary action principle). Hence, weak equivalence suffices for the existence of a hyper-regular Lagrangian.\footnote{Weak equivalence corresponds to a general treatment of Lagrangian mechanics that does not presuppose Riemannian geometry. As we discuss in Section~\ref{Curiel}, Equation~\ref{weak_equivalence} corresponds to the kinematical constraints central to Curiel's \parencites*[]{Curiel} argument for privileging Lagrangian mechanics.}

Whereas weak equivalence involves motion equivalence at the level of particle states, \textit{full equivalence} extends this to the entire distribution, i.e. the composite state of the system. Full equivalence holds when we can express the distribution---represented by the density---in terms of position and velocity variables. Formally, this requires that $\rho(q^i, p_j) = |J| \rho(x^i, v^j) = \left|\frac{\partial v^i}{\partial p_j}\right| \rho(x^i, v^j)$ since
\begin{equation}
\begin{aligned}
|J| &= \begin{vmatrix}
\frac{\partial x^i}{\partial q^j} & \frac{\partial x^i}{\partial p_j} \\
\frac{\partial v^i}{\partial q^j} & \frac{\partial v^i}{\partial p_j}
\end{vmatrix}
= \begin{vmatrix}
\delta^i_j & 0 \\
\frac{\partial v^i}{\partial q^j} & \frac{\partial v^i}{\partial p_j}
\end{vmatrix} \\
&= \left|\delta^i_j\right| \left|\frac{\partial v^i}{\partial p_j}\right| 
= \left|\frac{\partial v^i}{\partial p_j}\right|.
\end{aligned}
\end{equation}
Note that while the value given by $\rho(q^i, p_j)$ is invariant under unit changes, the value given by $\rho(x^i, v^j)$ depends on the choice of units through $\left|\frac{\partial v^i}{\partial p_j}\right|$. If $x^i$ truly sets the unit system by itself, then $\left|\frac{\partial v^i}{\partial p_j}\right|$ must be a function of position only, since the position variables---by themselves---define the units. Similar considerations hold for marginal distributions (i.e. distributions on a subset of the unit variables), entailing that all components of $\frac{\partial v^i}{\partial p_j}$ must be a function of position only. We set:\footnote{In Section~\ref{dependence}, we will see that $g^{ij}$ corresponds to the Riemannian metric, since $\partial v^i/\partial p_j$ it is a function only of position.}
\begin{equation}
\begin{aligned}
\frac{\partial v^i}{\partial p_j} = \frac{1}{m} g^{ij} \\
\frac{\partial p_j}{\partial v^i} = m g_{ji}
\end{aligned}
\end{equation}
where $m$ is the unit conversion constant between velocity and conjugate momentum, while $g_{ij}$ represents the linear dependency of the velocity on the momentum.
